Automatic prostate cancer detection through \acs*{dce}-\acs*{mri} images: all you need is a good normalization

### 2017-04-29 03_experiments/experiments.tex
Move to IEEE template

\ac{dce}-\ac{mri} is performed using a fat-suppressed 3D T$_1$ VIBE
sequence (\ac{tr}/\ac{te}/Flip angle:
\SI{3.25}{\ms}/\SI{1.12}{\ms}/\SI{10}{\degree}; Matrix: $256 \times
192$; \ac{fov}: $280 \times 210$ (with 75\% rectangular \ac{fov}); a slab of 16 partitions of \SI{3.5}{\mm} thickness; temporal resolution: \SI{6}{\s}/slab over approximately \SI{5}{\minute}).
A power injector (Medrad, Indianola, USA) is used to provide a bolus injection of Gd-DTPA (Dotarem, Guerbet, Roissy, France) at a dose of \SI{0.2}{\ml} Gd-DTPA/kg of body weight.

These \ac{dce}-\ac{mri} sequences are resampled using the spatial
information of the \ac{t2w}-\ac{mri} and any missing data are added using linear interpolation.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\citep{ibanez2005itk}.

@@

The implementation of the registration (C++), normalization (Python), and classification pipeline (Python) are publicly available on GitHub\footnote{\url{https://github.com/I2Cvb/lemaitre-2016-nov/tree/master}}~\citep{lemaitre2016github}.
The data used in this work are also publicly available\footnote{\url{https://zenodo.org/record/61163}}~\citep{lemaitre2016dce}.

@@

\begin{table*}
  \caption{Coefficient of determination $R^{2}$ (i.e., $\mu \ (\pm \sigma)$), while fitting data with the different quantification models.}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tabular}{lc c c c c c}
    \toprule
    Data type & Brix & Hoffmann & Tofts population \ac{aif} & Tofts patient \ac{aif} & \ac{pun} & Huisman \\
    \midrule
    Un-normalized & $0.85 \ (\pm 0.11)$ & $0.81 \ (\pm 0.17)$ & $0.84 \ (\pm 0.14)$ & $0.88 \ (\pm 0.12)$ & $0.27 \ (\pm 0.18)$ & $0.64 \ (\pm 0.24)$  \\
    Normalized    & $0.92 \ (\pm 0.05)$ & $0.72 \ (\pm 0.32)$ & $0.92 \ (\pm 0.06)$ & $0.90 \ (\pm 0.10)$ & $0.28 \ (\pm 0.20)$ & $0.75 \ (\pm 0.20)$  \\
    \bottomrule
  \end{tabular}
  }
  \label{tab:r2}
\end{table*}

@@

\begin{table*}
  \caption{\acs*{auc} (i.e., $\mu \ (\pm \sigma)$) for each individual pharmacokinetic parameter using a \acs*{rf} classifier.}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tabular}{l c c}
    \toprule
    \textbf{Features} & Un-normalized data & Normalized data \\
    \midrule
    \textbf{Brix model} & & \\
    \quad $A$         & $0.540\ (\pm 0.069)$ & $0.555\ (\pm 0.080)$ \\
    \quad $k_{el}$    & $0.549\ (\pm 0.062)$ & $0.577\ (\pm 0.093)$ \\
    \quad $k_{ep}$    & $0.506\ (\pm 0.032)$ & $0.497\ (\pm 0.019)$ \\
    \textbf{Hoffmann model} & & \\
    \quad $A$         & $0.516\ (\pm 0.020)$ & $0.508\ (\pm 0.031)$ \\
    \quad $k_{el}$    & $0.545\ (\pm 0.066)$ & $0.529\ (\pm 0.065)$ \\
    \quad $k_{ep}$    & $0.550\ (\pm 0.063)$ & $0.545\ (\pm 0.060)$ \\
    \textbf{Tofts model with population \ac{aif}} & & \\
    \quad $K_{trans}$ & $0.556\ (\pm 0.086)$ & $0.565\ (\pm 0.097)$ \\
    \quad $k_{ep}$    & $0.506\ (\pm 0.026)$ & $0.528\ (\pm 0.038)$ \\
    \quad $v_{p}$     & $0.533\ (\pm 0.064)$ & $0.548\ (\pm 0.082)$ \\
    \textbf{Tofts model with patient \ac{aif}} & & \\
    \quad $K_{trans}$ & $0.563\ (\pm 0.077)$ & $0.548\ (\pm 0.060)$ \\
    \quad $k_{ep}$    & $0.492\ (\pm 0.025)$ & $0.491\ (\pm 0.020)$ \\
    \quad $v_{p}$     & $0.530\ (\pm 0.069)$ & $0.495\ (\pm 0.033)$ \\
    \textbf{\ac{pun} model} & & \\
    \quad $a_0$       & $0.521\ (\pm 0.040)$ & $0.530\ (\pm 0.045)$ \\
    \quad $r$         & $0.550\ (\pm 0.085)$ & $0.573\ (\pm 0.097)$ \\
    \quad $\beta$     & $0.531\ (\pm 0.051)$ & $0.549\ (\pm 0.068)$ \\
    \textbf{Semi-quantitative analysis} & & \\
    \quad wash-in     & $0.587\ (\pm 0.107)$ & $0.533\ (\pm 0.032)$ \\
    \quad wash-out    & $0.516\ (\pm 0.037)$ & $0.486\ (\pm 0.035)$ \\
    \quad IAUC        & $0.506\ (\pm 0.048)$ & $0.513\ (\pm 0.032)$ \\
    \quad $\tau$      & $0.565\ (\pm 0.104)$ & $0.537\ (\pm 0.089)$ \\
    \quad $S_M - S_0$ & $0.560\ (\pm 0.083)$ & $0.532\ (\pm 0.029)$ \\
    \bottomrule
  \end{tabular}
  }
  \label{tab:resfeats}
\end{table*}

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfigure[Quantiative model without normalization.]{\label{fig:rfpharmaunorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/unormalized/unormalized_methods_0.pdf}} \hfill
  \subfigure[Quantitative model With normalization.]{\label{fig:rfpharmanorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/normalized/normalized_methods_0.pdf}}
  \hspace*{\fill}
  \hspace*{\fill}
  \subfigure[Entire enhanced \acs*{dce} signal classification.]{\label{fig:rfnormdcesignal}\includegraphics[width=.49\textwidth]{03_experiments/figures/full_signal_0.pdf}}
  \hspace*{\fill}
  \caption{\acs*{roc} analysis using a \acs*{rf} classifier with and
    without normalization using different approaches: \protect\subref{fig:rfpharmaunorm} -
    \protect\subref{fig:rfpharmanorm} pharmacokinetic and
    semi-quantitative models without and with normalization,
    respectively; \protect\subref{fig:rfnormdcesignal} entire enhanced
  \acs*{dce}-\acs*{mri} signal.}
  \label{fig:normpharmarf}
\end{figure*}

@@

To study the potential benefit of our normalization, \ac{cap} is detected at a voxel level using pharmacokinetic parameters estimated from non-normalized and normalized \ac{dce}-\ac{mri} data.
Each individual pharmacokinetic parameter is classified to evaluate its individual discriminative power to detect \ac{cap}.
In addition, p-values ($p$) are computed to identify pairs of classifiers
which are significantly different, using a Wilcoxon signed-ranked
test. Those p-values are represented in the color coded co-occurrence
matrix shown in Fig.\,\ref{fig:wilcoxon}, where blue cells correspond
to lower p-values while red cells are greater p-values.
Therefore, a \ac{rf} classifier is used in conjunction with a \ac{lopo}.
The use of \ac{rf} is encouraged since it leads to the best performance in the state-of-the-art methods~\citep{litjens2014computer,lemaitre2015computer}.
The results are summarized in Table~\ref{tab:resfeats} in terms of the \ac{auc}.
Normalization can improve the detection of \ac{cap}, but the benefit
of normalization is more obvious with the combination of the
pharmacokinetic features of a given model (e.g., $A$, $k_ep$, and
$k_el$ for the Brix model) as previously executed in the traditional \ac{cad} system~\citep{lemaitre2015computer}.
For the latter configuration, results are summarized by performing a \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
Quantification using normalized data outperforms quantification using
non-normalized data in terms of the classification performance, except
for Hoffmann and Tofts population-based \ac{aif} models.
The reasons behind the decrease of the \ac{auc} might be related to:
(i) a poor fitting as discussed in Sect.\,\ref{sec:fit} (cf., Hoffmann
model) and (ii) a small number of patients for the estimation of some parameters (cf., Tofts model).
The best classification performance is obtained using the
semi-quantitative approach with an \ac{auc} of $0.655 \pm 0.106$.

@@

\begin{figure*}
  \centering
  \includegraphics[width=1.\linewidth]{03_experiments/figures/plot_auc_by_zones.pdf}
  \caption{Comparison of the distribution of the \acs*{auc} score for
    each prostate zone, with and without normalization.}
  \label{fig:boxplot}
\end{figure*}

>>>>

\ac{dce}-\ac{mri} is performed using a fat-suppressed 3D T$_1$ VIBE
sequence (\ac{tr}/\ac{te}/Flip angle:
\SI{3.25}{\ms}/\SI{1.12}{\ms}/\SI{10}{\degree}; Matrix: $256 \times
192$; \ac{fov}: $280 \times 210$ (with 75\% rectangular \ac{fov}); a slab of 16 partitions of \SI{3.5}{\mm} thickness; temporal resolution: \SI{6}{\s}/slab over approximately \SI{5}{\minute}).
A power injector (Medrad, Indianola, USA) is used to provide a bolus
injection of Gd-DTPA (Dotarem, Guerbet, Roissy, France) at a dose of
\SI{0.2}{\ml} Gd-DTPA/kg of body weight.
The acquisition starts before the bolus injection to obtain
  pre-contrast volumes.

These \ac{dce}-\ac{mri} sequences are resampled using the spatial
information of the \ac{t2w}-\ac{mri} and any missing data are added using linear interpolation.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\cite{ibanez2005itk}.

@@

The implementation of the registration (C++), normalization (Python), and classification pipeline (Python) are publicly available on GitHub\footnote{\url{https://github.com/I2Cvb/lemaitre-2016-nov/tree/master}}~\cite{lemaitre2016github}.
The data used in this work are also publicly available\footnote{\url{https://zenodo.org/record/61163}}~\cite{lemaitre2016dce}.

@@

\begin{table*}
  \caption{Coefficient of determination $R^{2}$ (i.e., $\mu \ (\pm \sigma)$), while fitting data with the different quantification models.}
  \centering
  \begin{tabular}{lc c c c c c}
    \toprule
    Data type & Brix & Hoffmann & Tofts population \ac{aif} & Tofts patient \ac{aif} & \ac{pun} & Huisman \\
    \midrule
    Un-normalized & $0.85 \ (\pm 0.11)$ & $0.81 \ (\pm 0.17)$ & $0.84 \ (\pm 0.14)$ & $0.88 \ (\pm 0.12)$ & $0.27 \ (\pm 0.18)$ & $0.64 \ (\pm 0.24)$  \\
    Normalized    & $0.92 \ (\pm 0.05)$ & $0.72 \ (\pm 0.32)$ & $0.92 \ (\pm 0.06)$ & $0.90 \ (\pm 0.10)$ & $0.28 \ (\pm 0.20)$ & $0.75 \ (\pm 0.20)$  \\
    \bottomrule
  \end{tabular}
  \label{tab:r2}
\end{table*}

@@

\begin{table*}
  \caption{\acs*{auc} (i.e., $\mu \ (\pm \sigma)$) for each individual pharmacokinetic parameter using a \acs*{rf} classifier.}
  \centering
  \begin{tabular}{l c c}
    \toprule
    \textbf{Features} & Un-normalized data & Normalized data \\
    \midrule
    \textbf{Brix model} & & \\
    \quad $A$         & $0.540\ (\pm 0.069)$ & $0.555\ (\pm 0.080)$ \\
    \quad $k_{el}$    & $0.549\ (\pm 0.062)$ & $0.577\ (\pm 0.093)$ \\
    \quad $k_{ep}$    & $0.506\ (\pm 0.032)$ & $0.497\ (\pm 0.019)$ \\
    \textbf{Hoffmann model} & & \\
    \quad $A$         & $0.516\ (\pm 0.020)$ & $0.508\ (\pm 0.031)$ \\
    \quad $k_{el}$    & $0.545\ (\pm 0.066)$ & $0.529\ (\pm 0.065)$ \\
    \quad $k_{ep}$    & $0.550\ (\pm 0.063)$ & $0.545\ (\pm 0.060)$ \\
    \textbf{Tofts model with population \ac{aif}} & & \\
    \quad $K_{trans}$ & $0.556\ (\pm 0.086)$ & $0.565\ (\pm 0.097)$ \\
    \quad $k_{ep}$    & $0.506\ (\pm 0.026)$ & $0.528\ (\pm 0.038)$ \\
    \quad $v_{p}$     & $0.533\ (\pm 0.064)$ & $0.548\ (\pm 0.082)$ \\
    \textbf{Tofts model with patient \ac{aif}} & & \\
    \quad $K_{trans}$ & $0.563\ (\pm 0.077)$ & $0.548\ (\pm 0.060)$ \\
    \quad $k_{ep}$    & $0.492\ (\pm 0.025)$ & $0.491\ (\pm 0.020)$ \\
    \quad $v_{p}$     & $0.530\ (\pm 0.069)$ & $0.495\ (\pm 0.033)$ \\
    \textbf{\ac{pun} model} & & \\
    \quad $a_0$       & $0.521\ (\pm 0.040)$ & $0.530\ (\pm 0.045)$ \\
    \quad $r$         & $0.550\ (\pm 0.085)$ & $0.573\ (\pm 0.097)$ \\
    \quad $\beta$     & $0.531\ (\pm 0.051)$ & $0.549\ (\pm 0.068)$ \\
    \textbf{Semi-quantitative analysis} & & \\
    \quad wash-in     & $0.587\ (\pm 0.107)$ & $0.533\ (\pm 0.032)$ \\
    \quad wash-out    & $0.516\ (\pm 0.037)$ & $0.486\ (\pm 0.035)$ \\
    \quad IAUC        & $0.506\ (\pm 0.048)$ & $0.513\ (\pm 0.032)$ \\
    \quad $\tau$      & $0.565\ (\pm 0.104)$ & $0.537\ (\pm 0.089)$ \\
    \quad $S_M - S_0$ & $0.560\ (\pm 0.083)$ & $0.532\ (\pm 0.029)$ \\
    \bottomrule
  \end{tabular}
  \label{tab:resfeats}
\end{table*}

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfloat[Quantiative model without normalization.]{\label{fig:rfpharmaunorm}\includegraphics[width=.32\textwidth]{03_experiments/figures/unormalized/unormalized_methods_0.pdf}} \hfill
  \subfloat[Quantitative model With normalization.]{\label{fig:rfpharmanorm}\includegraphics[width=.32\textwidth]{03_experiments/figures/normalized/normalized_methods_0.pdf}}
  %\hspace*{\fill}
  %\hspace*{\fill}
  \hfill
  \subfloat[Entire enhanced \acs*{dce} signal classification.]{\label{fig:rfnormdcesignal}\includegraphics[width=.32\textwidth]{03_experiments/figures/full_signal_0.pdf}}
  \hspace*{\fill}
  \caption{\acs*{roc} analysis using a \acs*{rf} classifier with and
    without normalization using different approaches: \protect\subref{fig:rfpharmaunorm} -
    \protect\subref{fig:rfpharmanorm} pharmacokinetic and
    semi-quantitative models without and with normalization,
    respectively; \protect\subref{fig:rfnormdcesignal} entire enhanced
  \acs*{dce}-\acs*{mri} signal.}
  \label{fig:normpharmarf}
\end{figure*}

@@

To study the potential benefit of our normalization, \ac{cap} is detected at a voxel level using pharmacokinetic parameters estimated from non-normalized and normalized \ac{dce}-\ac{mri} data.
Each individual pharmacokinetic parameter is classified to evaluate its individual discriminative power to detect \ac{cap}.
In addition, p-values ($p$) are computed to identify pairs of classifiers
which are significantly different, using a Wilcoxon signed-ranked
test. Those p-values are represented in the color coded co-occurrence
matrix shown in Fig.\,\ref{fig:wilcoxon}, where blue cells correspond
to lower p-values while red cells are greater p-values.
Therefore, a \ac{rf} classifier is used in conjunction with a \ac{lopo}.
The use of \ac{rf} is encouraged since it leads to the best performance in the state-of-the-art methods~\cite{litjens2014computer,lemaitre2015computer}.
The results are summarized in Table~\ref{tab:resfeats} in terms of the \ac{auc}.
Normalization can improve the detection of \ac{cap}, but the benefit
of normalization is more obvious with the combination of the
pharmacokinetic features of a given model (e.g., $A$, $k_ep$, and
$k_el$ for the Brix model) as previously executed in the traditional \ac{cad} system~\cite{lemaitre2015computer}.
For the latter configuration, results are summarized by performing a \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
Quantification using normalized data outperforms quantification using
non-normalized data in terms of the classification performance, except
for Hoffmann and Tofts population-based \ac{aif} models.
The reasons behind the decrease of the \ac{auc} might be related to:
(i) a poor fitting as discussed in Sect.\,\ref{sec:fit} (cf., Hoffmann
model) and (ii) a small number of patients for the estimation of some parameters (cf., Tofts model).
The best classification performance is obtained using the
semi-quantitative approach with an \ac{auc} of $0.655 \pm 0.106$.

@@

\begin{figure}
  \centering
  \includegraphics[width=1.\linewidth]{03_experiments/figures/plot_auc_by_zones.pdf}
  \caption{Comparison of the distribution of the \acs*{auc} score for
    each prostate zone, with and without normalization.}
  \label{fig:boxplot}
\end{figure}
